Reproduce anaylses of Beninca et al (2008)

### 2016-02-11 code/report.Rmd
fixing zero removal issue

```{r, message=FALSE}
library(tidyr)
library(dplyr)
library(lubridate)
library(stringr)
library(ggplot2)
library(RCurl)
library(pracma)
library(oce)
library(tseriesChaos)
library(reshape2)
library(mgcv)
library(repmis)

@@

Format the dates as dates

```{r}
spp.abund$Date <- dmy(spp.abund$Date)
```

Ooops... R assumes the experiment was done in the 21st century. Shouldn't matter too much.

@@

```{r}
which(spp.abund$Date==ymd("2096-10-28 UTC"))
spp.abund$Date[701] <- ymd("2096-10-26 UTC")
```

Fix the Day.number problem:
```{r}
spp.abund$Day.number <- 1+as.numeric((spp.abund$Date - spp.abund$Date[1]) / 24 / 60 / 60)
```

@@

Now select only the date range used in the Nature paper. From the supplment *The analysis in Benincà et al. (Nature 2008) covered all data from 16/06/1991 until 20/10/1997*. (Remembering dates in the R dataframes are 2090s.)
```{r}
all.data <- filter(all.data, Date>=dmy("15/06/2091") & Date<=dmy("21/10/2097"))

@@

Aside: The ELE supplement contains the raw data and the transformed data, in separate data sheets. I (Owen) also got the interpolated data from Stephen Ellner directly.

## Interpolation

@@

Perform the interpolation:
```{r}
all.data1 <- na.omit(all.data)
mt <- plyr::dlply(all.data1,
                  "variable",
                  function(xx) pracma::interp1(x=xx$Day.number,
                                               y=xx$value,
                                               xi=xout,
                                               method="cubic"))
## Aside: the duplicated date that was previously fixed was only discovered by a warning message
## given by the pracma::interp1 function!

names(from.steve) <- c("Day.number", "variable", "value")

g1 <- ggplot(mt, aes(x=as.numeric(Day.number), y=value)) +
  facet_wrap(~variable, ncol=2, scales="free_y") +
  geom_line(size=0.5, col="black") +
  scale_colour_manual(values = species.colour.mapping)
g2 <- geom_line(data=from.steve, aes(x=Day.number, y=value), col="red")
g1 + g2

## Fourth root transform

@@

## Detrend

@@

@@

## Zero removal

> The time series of cyclopoid copepods, protozoa, filamentous diatoms, harpacticoid copepods and ostracods contained long sequences of zero values. This does not imply that these species were absent from the food web during these periods, but that their concentrations were below the detection limit. Time series dominated by many zeros can bias the statistical analysis. Therefore, these time series were shortened to remove long sequences of zero values, before the data transformation. The transformed data of all species in the food web are shown in Figure S2.

Since the method for this was not given, we do it by getting the trimmed day range from the fully transformed data in the ELE supplement.

```{r, echo=FALSE}

tr <- read.csv(text=getURL(paste0(repo,"/data/transformed_data_Nature2008.csv")),
               skip=7, na.string="")

tr <- tr[,-14:-24] ## remove bad columns
tr <- tr[-693:-694,] ## remove last two rows (contain summary stats)
tr <- gather(tr, key="Species", value="Abundance", 2:13)
levels(tr$Species)[levels(tr$Species)=="Calanoids"] <- "Calanoid.copepods"
levels(tr$Species)[levels(tr$Species)=="Total.Dissolved.Inorganic.Nitrogen"] <- "Total.dissolved.inorganic.nitrogen"
names(tr)[2] <- "variable"
names(tr)[1] <- "Day.number"
tr$Day.number <- as.numeric(as.character(tr$Day.number))
## get day range
min.max.day <- group_by(tr, variable) %>%
  summarise(min.day=min(Day.number[!is.na(Abundance)]), max.day=max(Day.number[!is.na(Abundance)]))

```{r}
trimmed_detrended <- full_join(detrended, min.max.day)
trimmed_detrended <- filter(trimmed_detrended, Day.number>min.day & Day.number<max.day)
```

## Rescale

@@

## Compare the data made above to published data

Now take a look at the transformed data provided in the ELE Supplement, in the data sheet *transformed_data_Nature2008*.

First note that the data in the ELE supplement include a data point at day 2658.1, whereas the data above finish at 2654.85. This is because the real data end at 2658, and the interpolation method above doesn't want to create data outside the range of the original data.

The graph isn't produced here, as there is a mismatch in days abundances were interpolated to. Should fix this...

@@

@@

two.datasets <- full_join(tr, scaled_trimmed_final)
#head(two.datasets)
#match(tr$Day.number, final$Day.number)

@@

ggplot(filter(two.datasets), aes(x=as.numeric(Day.number), y=stand.y)) +
  geom_path() +
  facet_wrap(~variable) +
  geom_line(aes(y=Abundance), col="red")

@@

ggplot(two.datasets, aes(x=Abundance, y=stand.y)) + geom_point() +
  facet_wrap(~variable)

@@

Looks OK, but suggests / shows that that data in the ELE supplement were standardised after removal of the zeros.

@@

ggplot(spec, aes(y=spec, x=1/freq, group=group)) + geom_line() + facet_wrap(~group, scales="free_y") + scale_x_log10() + 
coord_cartesian(ylim=c(0,40), xlim=c(10,240))

@@

#final_nozeros <- final[final$value!=0,]
final_nozeros <- scaled_trimmed_final

@@

qplot(x=as.vector(original.cors), y=as.vector(cor.coefs), ylim = c(-0.4,0.4), xlim=c(-0.4, 0.4),
      xlab="Correlations in original article", ylab="Correlations calculated in this reproduction") + geom_abline(intercept=0,slope=1)

@@

Then plot the graphs with LE:
```{r}
LEs <- mutate(LEs, days=20, Difference=-0.5)
g1 <- ggplot(diverg, aes(x=days, y=Difference)) + geom_point() + facet_wrap(~Species) +
  geom_text(data=LEs, aes(x=60, y=-2.5, label=round(le,3)), group=NULL)
g1
```

@@

@@

@@

Load and run the functions, or read in from data file (default option, for which you will need to change the path to the data file.)

>>>>

```{r, message=FALSE}
library(tidyr)
library(dplyr)
library(lubridate)
library(stringr)
library(ggplot2)
library(RCurl)
library(pracma)
library(oce)
library(tseriesChaos)
library(reshape2)
library(mgcv)
library(repmis)
library(magrittr)

@@

Format the dates as dates.

```{r}
spp.abund$Date <- dmy(spp.abund$Date)
# y123 <- year(spp.abund$Date) %% 100
# year(spp.abund$Date) <- 1900 + y123
# rm(y123)
```

@@

```{r}
which(spp.abund$Date==ymd("1996-10-28 UTC"))
spp.abund$Date[701] <- ymd("1996-10-26 UTC")
```

Check dates match the Day.number (should give false, as we changed a date):
```{r}
sum(spp.abund$Day.number == 1+as.numeric((spp.abund$Date - spp.abund$Date[1]) / 24 / 60 / 60)) == length(spp.abund$Date)
```

Recalculate the Day.number variable from the dates:
```{r}
spp.abund$Day.number <- 1+as.numeric((spp.abund$Date - spp.abund$Date[1]) / 24 / 60 / 60)
```

@@

Now select only the date range used in the Nature paper. From the supplment *The analysis in Benincà et al. (Nature 2008) covered all data from 16/06/1991 until 20/10/1997*. 
```{r}
all.data <- filter(all.data, Date>=dmy("15/06/1991") & Date<=dmy("21/10/1997"))

@@

The ELE supplement contains the raw data and the transformed data, in separate data sheets. I (Owen) also got the interpolated data from Stephen Ellner directly.

## Removal of long sequences of zeros

> The time series of cyclopoid copepods, protozoa, filamentous diatoms, harpacticoid copepods and ostracods contained long sequences of zero values. This does not imply that these species were absent from the food web during these periods, but that their concentrations were below the detection limit. Time series dominated by many zeros can bias the statistical analysis. Therefore, these time series were shortened to remove long sequences of zero values, before the data transformation.

The exact method for identifying the "long sequences of zeros" was not given. However, the transformed data with zeros removed was provided, and from this we can get the day range of the retained data:

```{r}
tr <- read.csv(text=getURL(paste0(repo,"/data/transformed_data_Nature2008.csv")),
               skip=7, na.string="")

tr <- tr[,-14:-24] ## remove bad columns
tr <- tr[-693:-694,] ## remove last two rows (contain summary stats)
tr <- gather(tr, key="Species", value="Abundance", 2:13)
levels(tr$Species)[levels(tr$Species)=="Calanoids"] <- "Calanoid.copepods"
levels(tr$Species)[levels(tr$Species)=="Total.Dissolved.Inorganic.Nitrogen"] <- "Total.dissolved.inorganic.nitrogen"
names(tr)[2] <- "variable"
names(tr)[1] <- "Day.number"
tr$Day.number <- as.numeric(as.character(tr$Day.number))
## get day range
min.max.day <- group_by(tr, variable) %>%
  summarise(min.day=min(Day.number[!is.na(Abundance)]), max.day=max(Day.number[!is.na(Abundance)]))
```

```{r}
trimmed_raw <- full_join(all.data, min.max.day)
trimmed_raw <- filter(trimmed_raw, Day.number>min.day & Day.number<max.day)
```

## Interpolation (zeros trimmed)

@@

Perform the interpolation:
```{r}

# make a function to do the interpolation
# next two lines make variables for testing
#x <- trimmed_raw %>% filter(variable=="Cyclopoids") %$% Day.number
#y <- trimmed_raw %>% filter(variable=="Cyclopoids") %$% value
Interpolate <- function(x, y) {
  xout.inner <- xout[xout>=min(x) & xout<=max(x)]
  res <- pracma::interp1(x=as.numeric(x),
                  y=as.numeric(y),
                  xi=xout.inner,
                  method="cubic")
  res.df <- data.frame(Day.number=xout.inner,
                       value=res)
}

# run the function and tidy the data
# (somewhat ashamed of this code...)

# remove NAs before interpolation, so they don't cause problems
trimmed_raw <- na.omit(trimmed_raw)
res <- group_by(trimmed_raw, variable) %>%
  do(le=Interpolate(.$Day.number, .$value))
for(p in 1:nrow(res)){
    temp <- data.frame(variable=res[p,1],
                       res[p,2][[1]])
    if(p==1)
      trimmed_interp <- temp
    if(p>1)
      trimmed_interp <- rbind(trimmed_interp, temp)
  }

#mt <- as.data.frame(mt)
#mt <- cbind(Day.number=xout, mt)
#mt <- gather(mt, variable, value, 2:13)
```

## Fourth root transform  (zeros trimmed)

@@

```{r}
trimmed_interp$fr.value <- trimmed_interp$value^0.25
```

## Detrend  (zeros trimmed)

@@

```{r}
ww.td <- filter(trimmed_interp, variable=="Total.dissolved.inorganic.nitrogen" |
                  variable=="Soluble.reactive.phosphorus" |
                  variable=="Bacteria" |
                  variable=="Ostracods" |
                  variable=="Harpacticoids")
## and to not detrend
ww.ntd <- filter(trimmed_interp, variable!="Total.dissolved.inorganic.nitrogen" &
                  variable!="Soluble.reactive.phosphorus" &
                  variable!="Bacteria" &
                  variable!="Ostracods" &
                  variable!="Harpacticoids")
## detrend:
ww1 <- group_by(ww.td, variable) %>%
  mutate(trend=ksmooth(Day.number,fr.value,bandwidth=300,kernel="normal")$y)
ww1$dt.value <- ww1$fr.value-ww1$trend
#ww1 <- select(ww1, trend)

@@

## Rescale (zeros trimmed)

@@

```{r}
## standardise
scaled_trimmed_final <- group_by(detrended, variable) %>%
  mutate(stand.y=as.numeric(scale(dt.value)))
summarise(scaled_trimmed_final, mean=mean(stand.y), sd=sd(stand.y))

## or don't standardise
#final <- detrended
#final$y <- final$dt.value
#summarise(final, mean=mean(y), sd=sd(y))
```

```{r}
#glimpse(scaled_trimmed_final)
```

## Compare the zeros removed transformed

@@

tr$Day.number <- as.character(tr$Day.number)
scaled_trimmed_final$Day.number <- as.character(scaled_trimmed_final$Day.number)

@@

two.datasets <- inner_join(tr, scaled_trimmed_final)
two.datasets$Day.number <- as.numeric(two.datasets$Day.number)

#head(two.datasets)
#match(tr$Day.number, final$Day.number)

@@

# ggplot(filter(two.datasets), aes(x=as.numeric(Day.number), y=stand.y)) +
#   geom_path() +
#   facet_wrap(~variable) +
#   geom_line(aes(y=Abundance), col="red")

# ggplot(filter(two.datasets, variable=="Soluble.reactive.phosphorus"), aes(x=as.numeric(Day.number), y=stand.y)) +
#   geom_path() +
#   facet_wrap(~variable) +
#   geom_line(aes(y=Abundance), col="red")

@@

#unique(two.datasets$variable)

two.datasets$variable <- plyr::revalue(two.datasets$variable,
                                 c(Soluble.reactive.phosphorus="P",
                                   Total.dissolved.inorganic.nitrogen="N"))

ggplot(two.datasets, aes(x=Abundance, y=stand.y)) + geom_point() +
  facet_wrap(~variable) +
  geom_abline(slope=1, intercept = 0, colour="red") +
  xlab("Values from published data") +
  ylab("Reproduced values")

@@

Looks pretty good, unclear why N and P are less well reproduced than others.

```{r}
final <- scaled_trimmed_final
final$y <- final$stand.y
```

@@

ggplot(spec, aes(y=spec, x=1/freq, group=group)) + geom_line() + facet_wrap(~group, scales="free_y") + scale_x_log10() + 
coord_cartesian(ylim=c(0,50), xlim=c(10,240))

@@

#final_nozeros <- final[final$value!=0,]
final_nozeros <- final

@@

qplot(x=as.vector(original.cors), y=as.vector(cor.coefs), ylim = c(-0.4,0.4), xlim=c(-0.4, 0.4),
      xlab="Correlations in original article", ylab="Correlations calculated in this reproduction") + geom_abline(intercept=0,slope=1, colour="red")

@@

Then plot the graphs with LE:
```{r}
LEs <- mutate(LEs, days=20, Difference=-0.5)

diverg$Species_short <- plyr::revalue(diverg$Species,
                                 c(Soluble.reactive.phosphorus="P",
                                   Total.dissolved.inorganic.nitrogen="N"))
LEs$Species_short <- plyr::revalue(LEs$Species,
                                 c(Soluble.reactive.phosphorus="P",
                                   Total.dissolved.inorganic.nitrogen="N"))

g1 <- ggplot(diverg, aes(x=days, y=Difference)) + geom_point() + facet_wrap(~Species_short) +
  geom_text(data=LEs, aes(x=60, y=0, label=round(le,3)))
g1
```

@@

Get the mean and standard deviation of the original and reproduced Lyapunov exponents

```{r}
mean(original.LEs$LE)
sd(original.LEs$LE)

mean(LEs$le)
sd(LEs$le)

@@

It seems that this must have been done on data without zeros removed...

## Interpolation (zeros retained)

Perform the interpolation:
```{r}
all.data <- na.omit(all.data)
mt <- plyr::dlply(all.data,
                  "variable",
                  function(xx) pracma::interp1(x=xx$Day.number,
                                               y=xx$value,
                                               xi=xout,
                                               method="cubic"))
## Aside: the duplicated date that was previously fixed was only discovered by a warning message
## given by the pracma::interp1 function!

names(from.steve) <- c("Day.number", "variable", "value")
from.steve$value_steve <- from.steve$value 
from.steve <- select(from.steve, -value)

## correcting an offset in the two datasets Day numbers
from.steve$Day.number <- from.steve$Day.number + 1.65 + 3.35

mt$Day.number <- as.character(mt$Day.number)
from.steve$Day.number <- as.character(from.steve$Day.number)
ff <- inner_join(mt, from.steve)
mt$Day.number <- as.numeric(mt$Day.number)
from.steve$Day.number <- as.numeric(from.steve$Day.number)
ff$Day.number <- as.numeric(ff$Day.number)

g1 <- ggplot(mt, aes(x=as.numeric(Day.number), y=sqrt(value))) +
  facet_wrap(~variable, ncol=2, scales="free_y") +
  geom_line(size=0.5, col="black") +
  scale_colour_manual(values = species.colour.mapping)
g2 <- geom_line(data=from.steve, aes(x=Day.number, y=sqrt(value_steve)), col="red")
g1 + g2

mt$Day.number <- as.character(mt$Day.number)
from.steve$Day.number <- as.character(from.steve$Day.number)
ff <- inner_join(mt, from.steve)
mt$Day.number <- as.numeric(mt$Day.number)
from.steve$Day.number <- as.numeric(from.steve$Day.number)
ff$Day.number <- as.numeric(ff$Day.number)

str(ff)

ff$difference <- ff$value - ff$value_steve
qplot(log10(ff$difference))

qplot(data=ff, x=log10(value_steve), y=log10(value)) +
  xlab("Values from Ellner (log 10 transformed)") +
  ylab("Reproduced values (log 10 transformed)") +
  geom_abline(intercept=0, slope=1, colour="red")

## Fourth root transform (zeros retained)

## Detrend (zeros retained)

## Rescale (zeros retained)

@@

Load and run the functions, or read in from data file.
